\Title

### 2012-12-20 synmut.tex
Methods

\section{Methods}
\comment{to be written\dots}
\section*{Acknowledgements}
\comment{to be written\dots}

>>>>

\section{Methods}
\subsection{Sequence data collection}
Longitudinal intrapatient viral RNA sequences were collected for published
studies~\citep{shankarappa_consistent_1999,
liu_selection_2006, bunnik_autologous_2008} and downloaded from the Los Alamos
National Laboratory (LANL) HIV sequence database~\citep{LANL2012}. The sequences from
some patients showed signs of HIV compartimentalization into subpopulations and
were discarded; a grand total of 11
patients with approximately 6 time points each and 10 sequences per time point
were analyzed. The time interval or resolution between two ocnsecutive sequences
was approximately 6 to 18 months.

\subsection{Sequence analisys}
The good sequences were aligned within each patient
via the translated amino acid sequence, using
Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence probed
by \citet{watts_architecture_2009}. Within each patient, a consensus RNA
sequence at the first time point was used to classify alleles as ancestral or
derived at all sites. Problematic sites that included large frequencies of gaps
were excluded from the analysis, because variable regions are known to be
subject to frequent indels, while our analysis is limited to nucleotide
substitutions. Time series of allele frequencies were extracted from the
sequences.

The synonymity of a mutation was assigned if the rest of the codon was
in the ancestral state and using the standard genetic code. Cases where more
than one mutation within the codon was observed were discarded. Slightly
different criteria for synonymous/nonsynonymous discrimination yielded similar
results.

\subsection{Fixation probability and secondary structure}
For the estimate of times to fixation/extinction, polymorphisms were
binned by frequency and the time to reaching the first boundary (fixation or
extinction) was stored. For the fixation probability, the long-time limit of the
resulting curves was used, excluding polymorphisms that arose late in the
clinical history (and would have had no time to reach either boundary).

For the correlation analysis with RNA secondary structure, the SHAPE scores were
downloaded from the journal website~\citep{watts_architecture_2009}. By virtue
of the alignment of the longitudinal sequences with the reference used by
\citet{watts_architecture_2009}, SHAPE reactivities were assigned to most sites.
Problematic assignments in indel-rich regions were excluded from the analysis.
In order to restrict the analysis to synonymous polymorphisms, a lower frequency
threshold of 0.15 was used (other thresholds yielded the same results). Since
very few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
enriched for to-be-lost mutations; hence the "lost" curve in \FIG{SHAPEA}
contains much more points than the "fixed" one.

The V loops and flanking regions were identified manually starting from the
annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
similar approach was used to label the C2-V5 region sequenced in
ref.~\citep{shankarappa_consistent_1999}.

\subsection{Computer simulations}
Simulations were performed using the recently published software
FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
\env{}-only simulations were performed and yielded comparable results. For each
set of parameters, approximately 100 simulation runs were averaged over. In each
run, a random fitness landscape with specified statistical properties (e.g.
density of beneficial sites, average deleterious effect of synonymous changes) was generated.
Although the curves shown in \FIG{simfixpvar} are not very smooth, small
parameter changes resulted in overall consistent trends across many repetitions.

For the discussion of simulation parameters, the areas below or above the neutral
diagonal were estimated from the binned fixation probabilities using the linear
interpolation between the bin centers. This measure is sufficiently precise for
our purposes, because the HIV data are quite scarse themselves.

\section*{Acknowledgements}
\comment{to be written\dots}
